---
title: "Analyzing the Accuracy of Google Maps Cycling Estimates"
output: html_document
---

I like riding my bike -- not only for exercise, but also for transportation whenever possible. I often use Google Maps to find a good route, but sometimes their cycling directions seem a bit off. I have always wondered how accurate these estimates actually are.

Recently, I have been experimenting with data provided by the [Bay Area Bike Share](http://www.bayareabikeshare.com/open-data) (BABS) system. I think bike sharing networks are a cool concept, so this has been a fun side project. This data set contains individual trip data complete with origin, destination, and trip duration, so it is perfect for investigating my question.

```{r echo=FALSE, eval=FALSE}
setwd("~/Documents/Projects/BikeShare/Rmd")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(stringr)
library(knitr)
library(ggplot2)
source("../R/helpers.R")
```

```{r warning=FALSE}
paths <- getFilePaths()
trip_data <- do.call("rbind", lapply(paths$trip, function(fn) {
  read_csv(fn) %>%
    rename(origin_id = `Start Terminal`,
           destination_id = `End Terminal`,
           actual_duration = Duration,
           trip_id = `Trip ID`) %>%
    mutate(time = as.POSIXct(`Start Date`, format = "%m/%d/%Y %H:%M"),
           year = format(time, "%Y"),
           month = format(time, "%m"),
           day = format(time, "%d"),
           hour = format(time, "%H"),
           minute = format(time, "%M"),
           weekday = ifelse(format(time, "%a") %in% c("Sat", "Sun"), "Weekend", "Weekday")) %>%
    select(trip_id, origin_id, destination_id, actual_duration, time, year, month, day, hour, minute, weekday)
}))
```

Through the Google Maps Directions API, I pull cycling estimates for every pair of stations in the same region.  See [here](https://github.com/llefebure/bike-sharing/blob/master/R/google-api-trip-estimates.R) for the full code that generates these numbers.

```{r}
gmaps_estimates <- readRDS("../data/gmaps_trip_estimates.Rds")
```

Next, I merge the trips data with the Google Maps estimates. Because some stations changed location, I need to make sure that each trip matches up with the correct estimate.

```{r}
trip_data_with_est <- inner_join(trip_data, gmaps_estimates) %>%
  mutate(origin_date_match = as.Date(time) >= start_date_origin & 
           as.Date(time) <= end_date_origin,
         destination_date_match = as.Date(time) >= start_date_destination & 
           as.Date(time) <= end_date_destination) %>%
  filter(is.na(origin_date_match) | origin_date_match == TRUE, 
         is.na(destination_date_match) | destination_date_match == TRUE) %>%
  group_by(trip_id) %>%
  arrange(origin_date_match, destination_date_match) %>%
  filter(row_number() == 1) %>%
  ungroup()
```

```{r}
summary(trip_data_with_est$actual_duration - trip_data_with_est$duration)
```

```{r}
ggplot(trip_data_with_est, aes(actual_duration - duration)) + geom_histogram() + xlim(c(-800, 800))
```